Information in recombination junctions can both compress and improve inferred tree sequences

### 2024-01-04 paper/writeup.tex
Some paper edits

\subsection{Algorithm}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)

\subsection*{Notation}
    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
     $N$ and $E$ are the sets of nodes and edges respectively,
     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome.
     The length of each tree $T_k$ is some subinterval $[a_k,a_{k+1})$ 
     with the lengths of consecutive trees in the sequence 
     only intersecting each intervals endpoints. 
    Additionally, a node $x\in N$ is \textit{unary} if it only has one child, 
    ie. only a single edge extending below it on the tree.
    With these tree sequences our goal is to
    minimize the number of edges
    as well as infer coalescence of genes. 
    To remove edges we consider the following example.
    Suppose for a tree $T_k$ in the tree sequence $TS$ 
    there exists a branch which contains a unary node $b\in N$.
    Then the node $b$ has one parent, $a\in N$, 
    and one child $c\in N$,
    each with edges $e_{ab}, e_{bc}\in E$ connecting them.
    If there is an edge between the nodes $a$ and $c$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $e_{ab},e_{bc}$ 
    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
    This action reduces the length of the edge $e_{ac}$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such unary nodes 
    over the entire tree sequence.

\begin{figure}
\begin{center}
\begin{tikzpicture} 
	\node[circle,draw,inner sep=1pt] (a) at (0,2) {a}
	[growth parent anchor=south] 
		child { node[circle,draw,inner sep=1pt] {b} 		
	 			child {
	 				node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
	 					child{node (d) at (0,1) {$T_k$}
	 						edge from parent [draw=none]}
	 				edge from parent 
	 					node[left] {$e_{b,c}$}
	 			}	 		
	 		edge from parent
	 			node[left] {$e_{a,b}$}
	 	};

	\node[circle,draw,inner sep=1pt] (a') at (3,2) {a}
	[growth parent anchor=south]
		child {
			child{
				node[circle,draw,inner sep=1pt] (c') at (0,-0.25) {c}	
					child{node (d) at (0,1) {$T_{k+1}$}
						edge from parent [draw=none]}
			}
			node[left] {$e_{a,c}$}
		};

	\node at (6,0) {$\Longrightarrow$};
	\node at (6,0.5) {\tn{Edge Extend}};

@@

\end{tikzpicture}
\end{center}
\caption{
    \label{fig:extending_diagram}
}
\end{figure}`

    In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
    be a tree sequence.
    Within tskit we can track a tree sequence using a function called `EdgeDiffs'.
    This function tracks edges which are removed and added 
    between consecutive trees in the sequence. 
    As in \ref{} (tikzfigure), we can track if 
    $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 
    when creating $T_{k+1}$, 
    and we can see if $e_{a,c}$ was just added in $T_{k+1}$.

>>>>

\subsection{Extend Edges}}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $TS$, 
    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
     $N$ and $E$ are the sets of nodes and edges respectively,
     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome.
     The length of each tree $T_k$ is some subinterval $[a_k,a_{k+1})$ 
     with the lengths of consecutive trees in the sequence 
     only intersecting each intervals endpoints.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying short paths (2 edges) in a tree.
    If neighboring trees contain edges with 
    identical parent and child nodes, we assume that there
    is implied coalescence of haplotypes between nodes, and
  	we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
  	In more concrete terms, for a tree $T_k$ in the tree sequence $TS$ 
    suppose there exists a path which contains a node $b\in N$,
    with parent and child nodes $a, c\in N$, respectively.
    If there is an edge between the nodes $a$ and $c$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $e_{ab},e_{bc}$ 
    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
    This action reduces the length of $e_{ac}$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such ancestral paths 
    over the entire tree sequence.

\begin{figure}[!ht]
\begin{center}
\begin{tikzpicture} 
	\node[circle,draw,inner sep=1pt] (a) at (0,2) {a}
	[growth parent anchor=south] 
		child { node[circle,draw,inner sep=1pt] {b} 		
	 			child {
	 				node[circle,draw,inner sep=1pt] (c) at (-0.5,0) {c}
	 				edge from parent 
	 					node[left] {$e_{b,c}$}}
 				child {
 					node (d) at (0,-1) {$T_k$}
 					edge from parent [draw=none]}
	 			child {
	 				node[circle,draw,inner sep=1pt] (e) at (0.5,0)  {d}
 					edge from parent
 						node[right] {$e_{b,d}$}} 		
	 		edge from parent
	 			node[left] {$e_{a,b}$}
	 	}; 	
	\node[circle,draw,inner sep=1pt] (a') at (4,2) {a}
	[growth parent anchor=south]
		child {
			child{
				node[circle,draw,inner sep=1pt] (c') at (0,-0.25) {c}	
					child{node (d) at (0,1) {$T_{k+1}$}
						edge from parent [draw=none]}
			}
			node[left] {$e_{a,c}$}
		};

	\node at (7,0) {$\Longrightarrow$};
	\node at (7,0.5) {\tn{Edge Extend}};

@@

\end{tikzpicture}
\end{center}
\caption{A simple visualization of the Edge Extend method.
    \label{fig:extending_diagram}
}
\end{figure}

    In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
    be a tree sequence.
    Within tskit we can track a tree sequence using a function called `EdgeDiffs'.
    This function tracks edges which are removed and added 
    between consecutive trees in the sequence. 
    As in \ref{fig:extending_diagram} (tikzfigure), we can track if 
    $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 
    when creating $T_{k+1}$, 
    and we can see if $e_{a,c}$ was just added in $T_{k+1}$.
